Financialisation of Commodity Markets

### 2024-10-02 docs/paper.qmd
First draft introduction

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon--commonly referred to as "financialisation"—and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. Using a bespoke asset pricing factors based framework we study the issue through the lens of co-movement and find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A vigorous legislative scrutiny accompanied by an equally intense policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section @sec-background for a detailed discussion.] eventually led to regulatory changes in several individual markets.  
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation--the burning issue of the day--at the individual commodity market level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.  
Financialisation was a phenomenon global in nature that, through the channel of index investment, affected the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems limited. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action.  Although appealing, for it comes with the promise of prompt "significant" results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of market fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure [@keynes_treatise_1930; @hicks_value_1939].  
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

We strive to provide one in this study by taking a fundamentally different empirical perspective that is broader in scope. We consider the entire cross-section of actively traded commodities on futures markets and use a futures-based asset pricing framework that includes factors constructed using both theoretical fundamentals and liquidity considerations.  

@@

By grouping together assets by shared attribute dynamics in common portfolios, asset pricing techniques--and factor models in particular--are well suited to study co-movement [@fama_common_1993; @carhart_persistence_1997; @asness_devil_2013; @fama_five_factor_2015; @hou_digesting_2015; @asness_fact_2015; @frazzini_betting_2014; @asness_quality_2019]. Careful, theory-based, attribute selection further enables a refined analysis of market fundamentals [@schwartz_short_2000; @miffre_momentum_2007; @gorton_fundamentals_2012; @cortazar_commodity_2013; @yang_investment_2013; @daskalaki_factors_2014; @szymanowska_anatomy_2014; @fernandez_skewness_2018; @bakshi_understanding_2019; @boons_basis_2019; @sakkas_factor_2020]. We therefore develop a futures-based asset pricing framework that includes factors constructed using both theoretical and liquidity considerations. Using the weekly returns, open interest and market position data described above we construct market portfolios of individual commodity nearby futures as well as mimicking portfolios for market, commercial hedging pressure, term structure and liquidity risk factors in returns. While our market portfolios are long only combinations of particular commodity sets, our other mimicking portfolios for risk factors are constructed as combinations of two sub-portfolios (legs), one held long, one held short, which respective constituents are pooled on a weekly basis according to risk factor specific criteria.  
All portfolios are equally weighted and returns for the two-legged factors are calculated as the difference between the average return for constituents of the long leg and that for the constituents of the short. i.e, return on the mimicking portfolio for risk factor $i$ ($f_{i}$) for period $t$:

@@

@tbl-stats-combined displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five--Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)--show a lower figure with five significant figures--Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)--against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  

>>>>

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. Using a bespoke asset pricing factors based framework we study the issue through the lens of co-movement and find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A rigorous legislative scrutiny accompanied by a vigorous policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section @sec-background for a detailed discussion.] eventually led to regulatory changes in several individual markets.  
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation---the burning issue of the day---at the individual commodity market level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.  
Financialisation was a phenomenon global in nature that, through the channel of index investment, affected the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems limited. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action.  Although appealing, for it comes with the promise of prompt "significant" results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure [@keynes_treatise_1930; @hicks_value_1939].  
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

We strive to provide one in this study by taking a fundamentally different empirical perspective that is broader in scope. We examine the interconnectedness of the commodity complex at the cross-sectional level, considering the entire cross-section of actively traded commodities on futures markets and use a bespoke futures-based asset pricing framework that includes factors constructed using both theoretical fundamentals and liquidity considerations.  

@@

By grouping together assets by shared attribute dynamics in common portfolios, asset pricing techniques---and factor models in particular---are well suited to study co-movement [@fama_common_1993; @carhart_persistence_1997; @asness_devil_2013; @fama_five_factor_2015; @hou_digesting_2015; @asness_fact_2015; @frazzini_betting_2014; @asness_quality_2019]. Careful, theory-based, attribute selection further enables a refined analysis of market fundamentals [@schwartz_short_2000; @miffre_momentum_2007; @gorton_fundamentals_2012; @cortazar_commodity_2013; @yang_investment_2013; @daskalaki_factors_2014; @szymanowska_anatomy_2014; @fernandez_skewness_2018; @bakshi_understanding_2019; @boons_basis_2019; @sakkas_factor_2020]. We therefore develop a futures-based asset pricing framework that includes factors constructed using both theoretical and liquidity considerations. Using the weekly returns, open interest and market position data described above we construct market portfolios of individual commodity nearby futures as well as mimicking portfolios for market, commercial hedging pressure, term structure and liquidity risk factors in returns. While our market portfolios are long only combinations of particular commodity sets, our other mimicking portfolios for risk factors are constructed as combinations of two sub-portfolios (legs), one held long, one held short, which respective constituents are pooled on a weekly basis according to risk factor specific criteria.  
All portfolios are equally weighted and returns for the two-legged factors are calculated as the difference between the average return for constituents of the long leg and that for the constituents of the short. i.e, return on the mimicking portfolio for risk factor $i$ ($f_{i}$) for period $t$:

@@

@tbl-stats-combined displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five---Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)---show a lower figure with five significant figures---Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)---against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  
